---
title: "Fundamentals of computational biology, fifth instalment"
author: "daniel.lundin@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 9
    fig_width: 8
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
---

```{r setup, echo=F, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = 'figures/', cache = TRUE)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries, message=F, cache = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
```

```{r constants}
```

```{r read-data}
# Read the ASV table and turn it long without zeroes
asvs <- read_tsv(
  '../../data/atacama-soils.asvtable.tsv',
  col_types = cols(.default = col_double(), seqid = col_character())
) %>%
  gather(sample, count, 2:67) %>%
  filter(count > 0)

# Read the taxonomy table
taxonomy <- read_tsv(
  '../../data/taxonomy.tsv',
  col_types = cols(
    .default = col_character(), Confidence = col_double()
  )
) %>%
  # Delete the D_0__ etc strings at the start of each taxon
  mutate(
    Taxon = gsub('D_[0-9]_+', '', Taxon)
  ) %>%
  # Rename Feature ID to seqid, the name in the ASV table
  rename(
    seqid = `Feature ID`,
  ) %>%
  # Separate th single taxonomy string into indvidiual taxa
  separate(
    Taxon, 
    c('domain', 'phylum', 'class', 'order', 'family', 'genus', 'species'), 
    sep = ';',
    fill = 'right'
  )
```

# Joins

Joining data from several tables is a fundamental operation in most analyses. In the Tidyverse, joins are performed
with the functions: `inner_join()`, `left_join()`, `right_join()` and `outer_join()`, that share the basic syntax
(see the builtin help or the relavant chapter in "R for data science"). In base R, the corresponding function is
`merge()`. As usual the Tidyverse functions are clearer and have fewer surprises.

All joins require at least one common variable to match data from two tables on. The difference between the four 
types is that an `inner_join()` will only return rows that were found in *both tables*, a `left_join()` will return
all rows from the first mentioned, *left* table but only the matching rows from the right table, a `right_join()` 
does the opposite to a `left_join()` and an `outer_join()` returns all rows from both tables. Most commonly used are
the `inner_join()` and the `left_join()`.

## "Natural join"

A natural join is a join that uses all shared variables between the two tables to join with. It does not require that
you specify by variables, but it's recommended.

```{r}
asvs %>% inner_join(taxonomy)
```

To get rid of the warning message -- 'Joining, by = "seqid"' -- you can specify that with `by`:

```{r}
asvs %>% inner_join(taxonomy, by = 'seqid')
```

If your key colums do not have the same name, you specify the join condition with `by` like this:

```{r}
# To create an example: Copy the taxonomy table but change the name of the seqid column to s, so that we can't do 
# a natural join
t <- taxonomy %>% rename(s = seqid)

# Now, you need to specify a by in the join
asvs %>% inner_join(t, by = c('seqid' = 's'))
```

If the *content* of the key columns are not exactly identical, you must reformat them so they're comparable:

```{r}
# To create an example: Modify the s column so it's not exactly the same but in a way that can be traced back
t <- t %>% mutate(s = sprintf("seqid:%s", s))

# A join now doesn't return anything, because there are no shared keys *contentwise*
asvs %>% inner_join(t, by = c('seqid' = 's'))

# You need to get rid of the "seqid:" before you can do the join
asvs %>% inner_join(t %>% mutate(s = sub('seqid:', '', s)), by = c('seqid' = 's'))
```

## `semi_join()` and `anti_join()`

`semi_join()` and `anti_join()` are tools when you want to use information in another table to filter a table
*without* adding the information from the second table to the first.

```{r}
# Returning only the Cyanobacteria
asvs %>% semi_join(taxonomy %>% filter(phylum == 'Cyanobacteria'), by = 'seqid')

# Returning everything except the Cyanobacteria with anti_join
asvs %>% anti_join(taxonomy %>% filter(phylum == 'Cyanobacteria'), by = 'seqid')
```

When you want to specify more than one phylum in the above example, use the `%in%` function:

```{r}
# Returning the Cyanobacteria and Firmicutes
asvs %>% semi_join(taxonomy %>% filter(phylum %in% c('Cyanobacteria', 'Firmicutes')), by = 'seqid')
```


